Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-13 article.tex
Modifikation des Teils von Marcus

\section{Extensions of dMod}
\label{sec:extensions}

>>>>

\section{Extensions of dMod}
\label{sec:extensions}
Computer algebra and symbolic tools are not part of \proglang{R}'s core functionality. In this section we illustrate two symbolic tools that are shiped with \pkg{dMod},  dealing with structural non-identifiability and steady-state constraints. They are implemented in \proglang{Python} and are interfaced via the \pkg{rPython} package.

\subsection{Lie-group symmetry detection}
In Section \ref{chap_uncertainty}, profile likelihood computation showed the existence of both practically and structurally non-identifiable parameters. While practical non-identifiability arises from insufficient information in the data, structural non-identifiabilty is connected to Lie-group symmetries, i.e. transformations of the states and parameters \[\Psi: (x,\theta) \longmapsto (x^*, \theta^*)\] that preserve the model prediction of the observables: \[g(x^*, \theta^*) = g(x,\theta)\,.\] Based on \citep{merkt2015higher}, the \code{symmetryDetection()} command outputs a list of available symmetry transformations. For example, the code
\begin{CodeChunk}
\begin{CodeInput}
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell", 
			 rate = "import_baso*TCA_buffer")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer", 
			 rate = "export_sinus*TCA_cell")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana", 
			 rate = "export_cana*TCA_cell")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer", 
			 rate = "reflux*TCA_cana")

observables <- eqnvec(
  buffer = "s*TCA_buffer",
  cellular = "s*(TCA_cana + TCA_cell)"
)

symmetryDetection(as.eqnvec(reactions), observables)
\end{CodeInput}
\end{CodeChunk}
returns the following output:
\begin{CodeChunk}
\begin{CodeInput}
1 transformation(s) found:
variable   : infinitesimal : transformation
-----------------------------------------------------
#1: Type: scaling
TCA_buffer : -TCA_buffer   : TCA_buffer*exp(-epsilon)
TCA_cana   : -TCA_cana     : TCA_cana*exp(-epsilon)
TCA_cell   : -TCA_cell     : TCA_cell*exp(-epsilon)
s          : s             : s*exp(epsilon)
\end{CodeInput}
\end{CodeChunk}

%outputs one scaling invariance for the parameters \code{s}, \code{TCA_buffer}, \code{TCA_cana} and \code{TCA_cell} which is 
In agreement with the identifiability analysis by the profile-likelihood method, the parameters \code{s}, \code{TCA_buffer}, \code{TCA_cana} and \code{TCA_cell} are found to be non-identifiable due to a scaling symmetry. The corresponding scaling transformation, last column, leaves the observation invariant for any choice of \code{epsilon}. The parameter non-identifiability can be resolved choosing one representative from the orbit of the transformation. In our case, the scaling parameter could e.g.~be fixed to 1.

%scaling invariance could be removed by fixing one of the parameters to a certain value unequal to zero\footnote{Fixing $\mbox{TCA\_buffer}=0$ does not remove the invariance, since the corresponding symmetry transformation is not valid \citep{merkt2015higher}.}. For the case of a translational invariance all values were allowed.

\subsection{Analytical steady-state constraints}
While for the the present model, the steady-state could be explicitly calculated by hand, this might be much more challenging for models with a large number of states and parameters. For many of these models, the \code{steadyStates()} command outputs an analytical steady-state solution that can be incorporated in the model as an additional parameter transformation. Based on \citep{rosenblatt2016customized}, the steady-state constraint is solved for a combination of state variables and kinetic parameters while positivity of the solution is ensured. As the paper states, the approach outperforms common methods of steady-state implementation with respect to reliability and performance of the optimization process. For our example, the code reads
\begin{CodeChunk}
\begin{CodeInput}
steadyStates(reactions, file = "SS.Rds)
\end{CodeInput}
\end{CodeChunk}
yielding the output

\begin{CodeChunk}
\begin{CodeInput}
TCA_cana   = TCA_buffer*export_cana*import_baso/(reflux*(export_cana + 
	     export_sinus))
TCA_cell   = TCA_buffer*import_baso/(export_cana + export_sinus)
TCA_buffer = TCA_buffer
\end{CodeInput}
\end{CodeChunk}

The solution is stored in an Rds file. After loading the file by \code{readRDS()}, the equations can be used when defining the parameter transformation.
